---
title: "Bayesian Networks in Formula One Race Strategy"
subtitle: "Round 1: Bahrain GP 2023"
author: "Natalia Kortals"
output: html_document
date: '2024-10-23'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(data.table)
library(GGally)
```

## EDA

Firstly, we load the data saved into .csv file, acquired using f1dataR package. https://CRAN.R-project.org/package=f1dataR

```{r load data}
BAH = data.frame(read.csv("data/bahrain2023.csv", header =TRUE))
head(BAH)
```

```{r}
summary(BAH)
```



Next, we plot the variables to understand the dataset: 

```{r plot pitstop vs laptime}
ggplot(data=subset(BAH, !is.na(pitstop_duration)), aes(x=pitstop_duration, y = lap_time)) +
  geom_point()  + geom_smooth(method=lm) +
  labs(x = "Pitstop Duration", y = "Lap Time", title = "Pitstop Duration vs Lap Times at Bahrain GP 2023") +
  theme(plot.title = element_text(hjust=0.5))
```


```{r plot team vs pitstop_duration}
# ggplot() + geom_boxplot(aes(x = BAH$team, y = BAH$pitstop_duration))

ggplot() + geom_boxplot(aes(x = BAH$team, y = BAH$pitstop_duration))+
  labs(x = "Team", y = "Pitstop Duration", title = "Box Plot of Pitstop Duration per Team (Bahrain 2023)") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r plot position vs laptime}
ggplot(data=BAH, aes(x=position, y = lap_time)) +
  geom_point()  + #geom_smooth(method=lm) +
  labs(x = "Position", y = "Lap Time", title = "Position vs Lap Times at Bahrain GP 2023") +
  theme(plot.title = element_text(hjust=0.5))
```

```{r plot team vs position}

ggplot() + geom_boxplot(aes(x = BAH$team, y = BAH$position))+
  labs(x = "Team", y = "Position", title = "Box Plot of Position per Team (Bahrain 2023)") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```
From the plot above, we can for example see that Red Bull Racing drivers stayed in the front the entire time, while Alpine drivers spanned positions between 16 and 10 - either because their drivers were far apart, or climbed or lost many positions throughout the event.

```{r}
BAH %>%
  ggplot( aes(x=lap_time, colour = tyre, fill = tyre)) +
    geom_density(alpha = 0.4)
```
```{r}
ggplot() + 
  geom_density(aes(x = BAH$lap_time), fill="#69b3a2", color="#e9ecef") + 
  labs(x = "Lap Time", y = "Density", title = "Density Plot of Lap Times at Bahrain GP 2023") +
  theme(plot.title = element_text(hjust=0.5))
```


```{r }
ggplot() + geom_boxplot(aes(x = BAH$team, y = BAH$lap_time))+
  labs(x = "Team", y = "Lap Time", title = "Box Plot of Lap Time per Team (Bahrain 2023)") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))

```

Before pairplot we need to set the categorical variables as factors

```{r}
BAH$driver <- factor(BAH$driver)
BAH$team <- factor(BAH$team)
BAH$tyre <- factor(BAH$tyre) 

# Circuit characteristics
BAH$traction <- factor(BAH$traction, ordered = TRUE)
BAH$asphalt_grip <- factor(BAH$asphalt_grip, ordered = TRUE)
BAH$asphalt_abrasion <- factor(BAH$asphalt_abrasion, ordered = TRUE)
BAH$track_evol <- factor(BAH$track_evol, ordered = TRUE)
BAH$tyre_stress <- factor(BAH$tyre_stress, ordered = TRUE)
BAH$braking <- factor(BAH$braking, ordered = TRUE)
BAH$lateral <- factor(BAH$lateral, ordered = TRUE)
BAH$downforce <- factor(BAH$downforce, ordered = TRUE)

```


```{r pairplot}
ggpairs(BAH, columns = c("lap_time", "team", "position", "pitstop_duration"))
```

```{r pistop density}
ggplot() + 
  geom_density(aes(x = BAH$pitstop_duration), fill="#ec8ccd", color="#ec8ccd") + 
  labs(x = "Pitstop Duration", y = "Density", title = "Density Plot of Pitstop Duration at Bahrain GP 2023") +
  theme(plot.title = element_text(hjust=0.5))
```

```{r pitstop binning}
BAH$pitstop_duration <- as.numeric(BAH$pitstop_duration)

BAH$pitstop_duration <- cut(BAH$pitstop_duration, breaks = c(min(BAH$pitstop_duration, na.rm = TRUE), quantile(BAH$pitstop_duration, probs = c(.2, .4,  .6, .8),na.rm=TRUE),max(BAH$pitstop_duration, na.rm = TRUE)), ordered_result = TRUE, include.lowest = TRUE)

```


```{r clean and order data}
BAH$driver <- factor(BAH$driver)
BAH$team <- factor(BAH$team)

# Circuit characteristics
BAH$traction <- factor(BAH$traction, ordered = TRUE)
BAH$asphalt_grip <- factor(BAH$asphalt_grip, ordered = TRUE)
BAH$asphalt_abrasion <- factor(BAH$asphalt_abrasion, ordered = TRUE)
BAH$track_evol <- factor(BAH$track_evol, ordered = TRUE)
BAH$tyre_stress <- factor(BAH$tyre_stress, ordered = TRUE)
BAH$braking <- factor(BAH$braking, ordered = TRUE)
BAH$lateral <- factor(BAH$lateral, ordered = TRUE)
BAH$downforce <- factor(BAH$downforce, ordered = TRUE)

# Lap time - ordered: the shorter the better
# the fastest lap is included thanks to argument include.lowest = TRUE
BAH$lap_time <- cut(BAH$lap_time, breaks = c(min(BAH$lap_time, na.rm = TRUE), quantile(BAH$lap_time,probs = c(.02, .05, .1, .15, .2, .3,.4,.5,.6, .7,.8, .85, .9, .95, .98),na.rm=TRUE),max(BAH$lap_time, na.rm = TRUE)), include.lowest = TRUE, ordered_result = TRUE)

# Add "no_pitstop" to the factor levels for pitstop_duration
BAH$pitstop_duration <- factor(BAH$pitstop_duration, 
                                levels = c("no_pitstop", levels(BAH$pitstop_duration)))
# Replace all NA values with "no_pitstop"
BAH <- BAH %>% replace_na(list(pitstop_duration = "no_pitstop"))


BAH$tyre <- factor(BAH$tyre)
BAH$rainfall <- factor(BAH$rainfall, ordered = TRUE)

# Make a variable for position on previous lap
BAH <- BAH %>%
  mutate(previous_position = ifelse(lap_number == 1, start_position, shift(BAH$position, n=1, type="lag")))

# Positions are ordered
BAH$previous_position <- factor(BAH$previous_position, ordered = TRUE)
BAH$position <- factor(BAH$position, ordered = TRUE)


```

```{r}
levels(BAH$lap_time)
```


```{r}
ggplot() + geom_bar(aes(x = BAH$lap_time))+
  labs(x = "Lap Time Intervals", y = "Count", title="Lap Time Binning Results for Bahrain GP 2023")+
theme(axis.text.x = element_text(angle = 45, hjust=1), plot.title = element_text(hjust=0.5))
```



```{r}
count_LTPT <- BAH %>%
  filter(pitstop_duration != "no_pitstop") %>%
  count(pitstop_duration, lap_time) %>%
  spread(lap_time, n, fill = 0)

heatmap_data <- gather(count_LTPT, key = "lap_time", value = "count", -pitstop_duration)

ggplot(heatmap_data, aes(x = lap_time, y = pitstop_duration, fill = count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Distribution of Pitstop Duration vs Lap Time",
       x = "Lap Time",
       y = "Pitstop Duration",
       fill = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))

```
```{r checking NAs}
# check which rows have NAs and investigate why
which(is.na(BAH))
```
```{r deal with NAs}
# dropping  nas
BAH <- drop_na(BAH)

```


```{r pitlane start warning}
# if starting position is 0, raise a Warning to check if that was a pitlane start
BAH %>%
  group_by(team) %>%
  summarize(has_zero = any(previous_position == 0), .groups = "drop") %>%
  filter(has_zero) %>%
  pull(team) -> drivers_with_zero

# Raise warning if any drivers seem to be starting from pitlane - check with reports on F1 website
if (length(drivers_with_zero) > 0) {
  warning(paste("The following drivers have starting position = 0:", paste(drivers_with_zero, collapse = ", ")))
}
```

# Simple DAG
## Known DAG.

Suppose we want to model the Bahrain GP with 5 variables: lap time, tyre compound, team, position, pitstop duration.


```{r}
library(bnlearn)
library(Rgraphviz)

# Choosing relevant variables
BAH1 <- BAH %>%  dplyr::select(c(lap_time, 
                                  tyre, team, position,
                                pitstop_duration)) 

dag = empty.graph(nodes = c("lap_time", "tyre", "team", "position", "pitstop_duration"))
# Known relationships
arc.set = matrix(c("team", "position",
                   "team", "tyre",
                   "lap_time", "position",
                   "tyre", "lap_time",
                  "pitstop_duration", "lap_time",
                   "pitstop_duration", "position",
                  "team", "pitstop_duration"), 
                 byrow = TRUE, ncol=2, dimnames =list(NULL,c("from","to")))
arcs(dag) = arc.set
graphviz.plot(dag)
```
```{r}
arc.strength(dag, data = BAH1, criterion = "x2")
```


```{r}
bn.bayes = bn.fit(dag,data=BAH1, method = "bayes", iss = 10)

bn.bayes$pitstop_duration
```

```{r}
bn.fit.barchart(bn.bayes$lap_time)
```

```{r}
bn.fit.barchart(bn.bayes$tyre)
```

```{r}
bn.bayes$lap_time
```

```{r}
bn.bayes$tyre
```
## DAG unknown: constraint-based method (incremental association Markov blanket)

```{r}
BAH.iamb_naive =iamb(BAH1, test ='mi')
graphviz.plot(BAH.iamb_naive)
```

Make a blacklist representing the natural temporal ordering of variables

```{r}
# Blacklist variables so that chronological order is kept
bl = matrix(c(
              "lap_time", "pitstop_duration",
              "lap_time", "tyre"
             ), 
            ncol = 2, byrow = TRUE)

# Nothing from position
for (name in names(BAH1)){
  bl2 = matrix(c("position", name), ncol=2,
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl2)
}


# Nothing to team
for (name in names(BAH1)){
  bl3 = matrix(c(name, "team"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl3)
}



bl
```




```{r}
BAH.iamb =iamb(BAH1, blacklist = bl, test ='mi')
graphviz.plot(BAH.iamb)
```

### Perform independence tests



```{r}
# Does pitstop influence lap time?
ci.test("lap_time","pitstop_duration", test = "mc-jt", data = BAH1)
```
```{r}
# Does team influence position?
ci.test("position", "team", test = "mc-mi", data = BAH1)
```
```{r}
# Does team influence tyre?
ci.test("tyre", "team", test = "mc-mi", data = BAH1)
```

```{r}
# Does team influence pitstop_duration?
ci.test("pitstop_duration", "team", test = "mc-mi", data = BAH1)
```
```{r}
# Does team influence lap_time?
ci.test("lap_time", "team", "pitstop_duration", test = "mc-mi", data = BAH1)
```

```{r}
# Does lap_time influence position?
ci.test("position", "lap_time", "team", test = "mc-mi", data = BAH1)
```


```{r}
# Add the arcs from conditional independence tests with resulting p-value < 0.05
tested_BAH.iamb = set.arc(BAH.iamb, from = "lap_time", to = "position")
tested_BAH.iamb = set.arc(tested_BAH.iamb, from = "team", to = "lap_time")
tested_BAH.iamb = set.arc(tested_BAH.iamb, from = "team", to = "pitstop_duration")


graphviz.plot(tested_BAH.iamb)
```
```{r}
# Does pitstop influence position?
ci.test("position", "pitstop_duration", "team", test = "mc-mi", data = BAH1)
```



```{r}
# Does tyre influence lap time?
ci.test("lap_time","tyre", c("pitstop_duration"), test = "mc-mi", data = BAH1)
```
```{r}
tested_BAH.iamb = set.arc(tested_BAH.iamb, from = "pitstop_duration", to = "position")
tested_BAH.iamb = set.arc(tested_BAH.iamb, from = "tyre", to = "lap_time")
graphviz.plot(tested_BAH.iamb)
```


## DAG uknown: score-based method (hill-climbing)

```{r}
# Use hill-climbing to determine the DAG
BAH.hc1 =hc(BAH1)
graphviz.plot(BAH.hc1)
```

Learning DAG using hill climbing and whitelists, blacklists


```{r}
# Use hill-climbing to determine the DAG with blacklisting
BAH.hc_bl =hc(BAH1, blacklist = bl)
graphviz.plot(BAH.hc_bl)
```
```{r}
arc.strength(BAH.hc_bl, data = BAH1, criterion = "x2")
```

### Scores

```{r}
# Known
score(dag, data = BAH1, type = "bic")

# IAMB 
score(BAH.iamb, data = BAH1, type = "bic")
score(tested_BAH.iamb, data = BAH1, type = "bic")

# HC
score(BAH.hc1, data = BAH1, type = "bic")
score(BAH.hc_bl, data = BAH1, type = "bic")

```


Here we can note that the CPDAG underlying hc and hc_bl is the same, so their dags are score-equivalent. Indeed, we get the same BIC score for both.

# Dynamic DAG
## Known DAG.

Suppose we want to model the Bahrain GP with 5 variables: lap time, tyre compound, team, position, pitstop duration.


```{r}
library(bnlearn)
library(Rgraphviz)

# Choosing relevant variables
BAH2 <- BAH %>%  dplyr::select(c(lap_time, previous_position,
                                  tyre, team, position,
                                pitstop_duration)) 

dag_d = empty.graph(nodes = c("lap_time", "tyre", "team", "position", "pitstop_duration", "previous_position"))
# Known relationships
arc.set = matrix(c("team", "position",
                   "team", "tyre",
                   "lap_time", "position",
                   "tyre", "lap_time",
                  "pitstop_duration", "lap_time",
                   "pitstop_duration", "position",
                  "team", "pitstop_duration",
                  "previous_position", "position"), 
                 byrow = TRUE, ncol=2, dimnames =list(NULL,c("from","to")))
arcs(dag_d) = arc.set
graphviz.plot(dag_d)
```
```{r}
arc.strength(dag_d, data = BAH2, criterion = "x2")
```


```{r}
bn.bayes_d = bn.fit(dag_d,data=BAH2, method = "bayes", iss = 10)

bn.bayes_d$pitstop_duration
```

```{r}
bn.fit.barchart(bn.bayes_d$lap_time)
```

```{r}
bn.fit.barchart(bn.bayes_d$tyre)
```

```{r}
bn.bayes_d$lap_time
```

```{r}
bn.bayes_d$tyre
```
## DAG unknown: constraint-based method (incremental association Markov blanket)

```{r}
BAH.iamb_naive_d =iamb(BAH2, test ='mi')
graphviz.plot(BAH.iamb_naive_d)
```

Make a blacklist representing the natural temporal ordering of variables

```{r}
# Blacklist variables so that chronological order is kept
bl = matrix(c(
              "lap_time", "pitstop_duration",
              "lap_time", "tyre"
             ), 
            ncol = 2, byrow = TRUE)

# Nothing from position
for (name in names(BAH2)){
  bl2 = matrix(c("position", name), ncol=2,
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl2)
}


# Nothing to team
for (name in names(BAH2)){
  bl3 = matrix(c(name, "team"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl3)
}

bl

wl = matrix(c(
              "previous_position", "position"
             ), 
            ncol = 2, byrow = TRUE)
```




```{r}
BAH.iamb_d =iamb(BAH2, blacklist = bl, whitelist = wl, test ='mi')
graphviz.plot(BAH.iamb_d)
```

### Perform independence tests



```{r}
# Does pitstop influence lap time?
ci.test("lap_time","pitstop_duration", test = "mc-jt", data = BAH2)
```
```{r}
# Does team influence position?
ci.test("position", "team", test = "mc-mi", data = BAH2)
```
```{r}
# Does team influence tyre?
ci.test("tyre", "team", test = "mc-mi", data = BAH2)
```

```{r}
# Does team influence pitstop_duration?
ci.test("pitstop_duration", "team", test = "mc-mi", data = BAH2)
```
```{r}
# Does team influence lap_time?
ci.test("lap_time", "team", "pitstop_duration", test = "mc-mi", data = BAH2)
```

```{r}
# Does lap_time influence position?
ci.test("position", "lap_time", "team", test = "mc-mi", data = BAH2)
```


```{r}
# Add the arcs from conditional independence tests with resulting p-value < 0.05
tested_BAH.iamb_d = set.arc(BAH.iamb_d, from = "lap_time", to = "position")
tested_BAH.iamb_d = set.arc(tested_BAH.iamb_d, from = "team", to = "lap_time")
tested_BAH.iamb_d = set.arc(tested_BAH.iamb_d, from = "team", to = "pitstop_duration")


graphviz.plot(tested_BAH.iamb_d)
```
```{r}
# Does pitstop influence position?
ci.test("position", "pitstop_duration", c("lap_time", "previous_position"), test = "mc-jt", data = BAH2)
```



```{r}
# Does tyre influence lap time?
ci.test("lap_time","tyre", c("pitstop_duration", "team"), test = "mc-mi", data = BAH2)
```

```{r}
# Does team influence previous_position?
ci.test("previous_position","team", test = "mc-mi", data = BAH2)
```

```{r}
# Does team influence position?
ci.test("position","team", "previous_position", test = "mc-mi", data = BAH2)
```

```{r}
tested_BAH.iamb_d = set.arc(tested_BAH.iamb_d, from = "tyre", to = "lap_time")
tested_BAH.iamb_d = set.arc(tested_BAH.iamb_d, from = "team", to = "position")
tested_BAH.iamb_d = set.arc(tested_BAH.iamb_d, from = "team", to = "previous_position")
graphviz.plot(tested_BAH.iamb_d)
```


## DAG uknown: score-based method (hill-climbing)

```{r}
# Use hill-climbing to determine the DAG
BAH.hc2 =hc(BAH2)
graphviz.plot(BAH.hc2)
```

Learning DAG using hill climbing and whitelists, blacklists


```{r}
# Use hill-climbing to determine the DAG with blacklisting
BAH.hc_bl_d =hc(BAH2, blacklist = bl, whitelist = wl)
graphviz.plot(BAH.hc_bl_d)
```
```{r}
arc.strength(BAH.hc_bl_d, data = BAH2, criterion = "x2")
```

### Scores

```{r}
# Known
score(dag_d, data = BAH2, type = "bic")

# IAMB 
score(BAH.iamb_d, data = BAH2, type = "bic")
score(tested_BAH.iamb_d, data = BAH2, type = "bic")

# HC
score(BAH.hc2, data = BAH2, type = "bic")
score(BAH.hc_bl_d, data = BAH2, type = "bic")

```

The known DAG seems to heavily underrepresent the dataset, with a very low BIC score. Notably, the IAMB with blacklisting fits the data better before the addition of arcs from conditional independence tests. Hill climbing results suggest that the addition of blacklists has not affected the score by much, with the penalty of adding more arcs.
