---
title: "Bayesian Networks in Formula One Race Strategy"
subtitle: "Round 1: Bahrain GP 2023"
author: "Natalia Kortals"
output: html_document
date: '2024-10-23'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(data.table)
library(GGally)
```

Firstly, we load the data saved into .csv file, acquired using f1dataR package. https://CRAN.R-project.org/package=f1dataR

```{r load data}
BAH = data.frame(read.csv("data/bahrain2023.csv", header =TRUE))
head(BAH)
```

```{r}
summary(BAH)
```

Before pairplot we need to set the categorical variables as factors


```{r}
BAH$driver <- factor(BAH$driver)
BAH$team <- factor(BAH$team)

# Circuit characteristics
BAH$traction <- factor(BAH$traction, ordered = TRUE)
BAH$asphalt_grip <- factor(BAH$asphalt_grip, ordered = TRUE)
BAH$asphalt_abrasion <- factor(BAH$asphalt_abrasion, ordered = TRUE)
BAH$track_evol <- factor(BAH$track_evol, ordered = TRUE)
BAH$tyre_stress <- factor(BAH$tyre_stress, ordered = TRUE)
BAH$braking <- factor(BAH$braking, ordered = TRUE)
BAH$lateral <- factor(BAH$lateral, ordered = TRUE)
BAH$downforce <- factor(BAH$downforce, ordered = TRUE)

BAH$tyre <- factor(BAH$tyre, ordered = TRUE
                   # , levels = c("SOFT", "MEDIUM", "HARD")
                   ) 
```


```{r pairplot}


ggpairs(BAH, columns = c("lap_time", "team", "position", "pitstop_duration")
        # ,aes(color = tyre, alpha = 0.5)
             )
```

Todo:
plot laptime vs position scatter, maybe coloured by tyre
team vs position boxplot
team vs laptime boxplot, with colour for tyres too
tyre vs laptime density plots, coloured by tyre overlayed on top of each other
pitstop_duration vs team





Next, we plot the variables to understand the dataset: 

```{r plot pitstop vs laptime}
ggplot(data=subset(BAH, !is.na(pitstop_duration)), aes(x=pitstop_duration, y = lap_time)) +
  geom_point()  + geom_smooth(method=lm)
```

```{r plot tyre count}
ggplot() + geom_bar(aes(x = BAH$tyre))
```

```{r plot team vs pitstop_duration}
ggplot() + geom_boxplot(aes(x = BAH$team, y = BAH$pitstop_duration))
```
```{r tyre vs laptime}
ggplot() + geom_boxplot(aes(x = BAH$tyre, y = BAH$lap_time))
```




```{r}
BAH %>%
  ggplot( aes(x=lap_time, colour = tyre, fill = tyre)) +
    geom_density(alpha = 0.4)
```
```{r}
ggplot() + 
  geom_density(aes(x = BAH$lap_time), fill="#69b3a2", color="#e9ecef") + 
  labs(x = "Lap Time", y = "Density", title = "Density Plot of Lap Times at Bahrain GP 2023") +
  theme(plot.title = element_text(hjust=0.5))
```


```{r }
ggplot() + geom_point(aes(x = BAH$team, y = BAH$lap_time, colour = BAH$tyre))

```

```{r}
min(BAH$lap_time, na.rm = TRUE)
```

```{r finding appropriate pitstop bins}
#Reload the unprocessed dataset
BAH = data.frame(read.csv("data/bahrain2023.csv", header =TRUE))
```

```{r}
ggplot() + 
  geom_density(aes(x = BAH$pitstop_duration), fill="#ec8ccd", color="#ec8ccd") + 
  labs(x = "Pitstop Duration", y = "Density", title = "Density Plot of Pitstop Duration at Bahrain GP 2023") +
  theme(plot.title = element_text(hjust=0.5))
```

```{r}
# Convert pitstop_duration to numeric
BAH$pitstop_duration <- as.numeric(BAH$pitstop_duration)
# Make bins
BAH$pitstop_duration <- cut(BAH$pitstop_duration, breaks = c(min(BAH$pitstop_duration, na.rm = TRUE), quantile(BAH$pitstop_duration, probs = c(.2, .4,  .6, .8),na.rm=TRUE),max(BAH$pitstop_duration, na.rm = TRUE)), ordered_result = TRUE, include.lowest = TRUE, labels = c("fastest_pit", ".2", ".4", ".6", ".8"))

ggplot(data=subset(BAH, !is.na(pitstop_duration)), aes(x=pitstop_duration)) +
  geom_bar()
```

```{r}
ggplot() + geom_point(aes(x = BAH$pitstop_duration, y = BAH$lap_time))
```


```{r clean and order data}
BAH$driver <- factor(BAH$driver)
BAH$team <- factor(BAH$team)

# Circuit characteristics
BAH$traction <- factor(BAH$traction, ordered = TRUE)
BAH$asphalt_grip <- factor(BAH$asphalt_grip, ordered = TRUE)
BAH$asphalt_abrasion <- factor(BAH$asphalt_abrasion, ordered = TRUE)
BAH$track_evol <- factor(BAH$track_evol, ordered = TRUE)
BAH$tyre_stress <- factor(BAH$tyre_stress, ordered = TRUE)
BAH$braking <- factor(BAH$braking, ordered = TRUE)
BAH$lateral <- factor(BAH$lateral, ordered = TRUE)
BAH$downforce <- factor(BAH$downforce, ordered = TRUE)

# Lap time - ordered: the shorter the better
# the fastest lap is included thanks to argument include.lowest = TRUE
BAH$lap_time <- cut(BAH$lap_time, breaks = c(min(BAH$lap_time, na.rm = TRUE), quantile(BAH$lap_time,probs = c(.02, .05, .1, .15, .2, .3,.4,.5,.6, .7,.8, .85, .9, .95, .98),na.rm=TRUE),max(BAH$lap_time, na.rm = TRUE)), include.lowest = TRUE, ordered_result = TRUE
                    # without the labels yet
                    # , labels = c("fastest", ".05", ".1", ".15", ".2", ".25", ".3",".4",".5",".6", ".7",".8", ".85", ".87", ".9", ".92", ".94", ".96", ".98")
                    )

# Pitstop duration bin and fill NAs with a [0,1] bin
levels(BAH$pitstop_duration) <- c("[0,1]", levels(BAH$pitstop_duration))
BAH <- BAH %>% replace_na(list(pitstop_duration = "[0,1]")) #change to "NA"


BAH$tyre <- factor(BAH$tyre)
BAH$rainfall <- factor(BAH$rainfall, ordered = TRUE)

# Make a variable for position on previous lap
BAH <- BAH %>%
  mutate(previous_position = ifelse(lap_number == 1, start_position, shift(BAH$position, n=1, type="lag")))

# Positions are ordered
BAH$previous_position <- factor(BAH$previous_position, ordered = TRUE)
BAH$position <- factor(BAH$position, ordered = TRUE)

# Choosing relevant variables
BAH <- BAH %>%  dplyr::select(c(lap_time, 
                                  tyre, team, position, 
                                  previous_position, pitstop_duration, driver
                                # traction, asphalt_grip, asphalt_abrasion, 
                                # track_evol, tyre_stress, braking, 
                                # lateral, downforce
                                )) #  rainfall, driver

```



```{r checking NAs}
# check which rows have NAs and why
which(is.na(BAH))
```
```{r deal with NAs}
# dropping  nas
BAH <- drop_na(BAH)
BAHapest <- BAH
```
```{r}
ggplot() + geom_bar(aes(x = BAH$lap_time))+
  labs(x = "Lap Time Intervals", y = "Count", title="Lap Time Binning Results for Bahrain GP 2023")+
theme(axis.text.x = element_text(angle = 45, hjust=1), plot.title = element_text(hjust=0.5))
```




```{r}
ggplot() + geom_point(aes(x = BAH$pitstop_duration, y = BAH$lap_time))
```
```{r}
ggplot() + geom_point(aes(x = BAH$team, y = BAH$pitstop_duration))
```

```{r}
# if starting position is 0, raise a Warning to check if that was a pitlane start
# Group by 'driver' and check for zeros
BAH %>%
  group_by(team) %>%
  summarize(has_zero = any(previous_position == 0), .groups = "drop") %>%
  filter(has_zero) %>%
  pull(team) -> drivers_with_zero

# Raise warning if any drivers have zeros
if (length(drivers_with_zero) > 0) {
  warning(paste("The following drivers have zeros in their values:", paste(drivers_with_zero, collapse = ", ")))
}
```

## Starting with a know DAG.

```{r}
library(bnlearn)
library(Rgraphviz)
dag = empty.graph(nodes = c("lap_time", "tyre", "team", "position", 
                      "previous_position", "pitstop_duration")) # "rainfall",
arc.set = matrix(c("team", "position",
                   "lap_time", "position",
                   "tyre", "lap_time",
                  "pitstop_duration", "lap_time",
                   "pitstop_duration", "position",
                  # "rainfall", "lap_time",
                  # "rainfall", "tyre",
                  "team", "pitstop_duration",
                  "previous_position", "position"
                  
                  ), 
                 byrow = TRUE, ncol=2, dimnames =list(NULL,c("from","to")))
arcs(dag) = arc.set
graphviz.plot(dag)
```
```{r}
arc.strength(dag, data = BAH, criterion = "x2")
```


```{r}
bn.bayes = bn.fit(dag,data=BAH, method = "bayes", iss = 10)

bn.bayes$pitstop_duration
```
```{r}
bn.bayes$lap_time
```

```{r}
bn.bayes$tyre
```
## DAG uknown case

Learning DAG using hill climbing and  blacklists

```{r}
# Blacklist variables so that chronological order is kept
bl = matrix(c("position", "previous_position",
              "pitstop_duration", "tyre",
              "tyre", "pitstop_duration",
              "lap_time", "tyre"
              
              ), 
            ncol = 2, byrow = TRUE)

# Nothing from position
for (name in names(BAH)){
  bl2 = matrix(c("position", name), ncol=2,
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl2)
}


# Nothing to previous position apart from team
for (name in names(BAH)){
  if (name != "team"){
  bl3 = matrix(c(name, "previous_position"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl3)}
}

# Nothing to team
for (name in names(BAH)){
  bl4 = matrix(c(name, "team"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl4)
}

bl
```


```{r}
unknown_dag = empty.graph(nodes = names(BAH))
wl = matrix(c("previous_position", "position",
              "lap_time", "position"
              ), 
            ncol = 2, byrow = TRUE)
arcs(unknown_dag) = wl
graphviz.plot(unknown_dag)

```

There is a lot of blacklisting/ whitelisting involved, which pretty much determines the DAG - is this an indication it's better to use a known DAG with distributions of variables?

```{r}
# Use hill-climbing to determine the DAG
BAH.hc =hc(BAH, blacklist = bl, whitelist = wl)
graphviz.plot(BAH.hc)
```

```{r}
arc.strength(BAH.hc, data = BAH, criterion = "x2")
```

```{r}
score(dag, data = BAH, type = "bic")
score(BAH.hc, data = BAH, type = "bic")
```
```{r}
score(dag, data = BAH, type = "bde")
score(BAH.hc, data = BAH, type = "bde")
```

```{r}
# Alt hill cclimbing with no rules
altBAH.hc =hc(BAH, blacklist = bl)
score(altBAH.hc, data = BAH, type = "bde")
graphviz.plot(altBAH.hc)
```

## Use incremental association Markov blanket to learn the DAG

```{r}
BAH.iamb =iamb(BAH, blacklist = bl, whitelist = wl, test ='mi')
graphviz.plot(BAH.iamb)
```


## Perform independence tests

```{r}
# Testing for conditional independence: position
ci.test("team", "previous_position", test = "mi", data = BAH)
```
```{r}
dag.BAH = set.arc(BAH.iamb, from = "team", to = "previous_position")

ci.test("team", "pitstop_duration", test = "mi", data = BAH)
```
```{r}
ci.test("lap_time","pitstop_duration", test = "jt", data = BAH)
```

```{r}
ci.test("lap_time", "team", "tyre", test = "mi", data = BAH)
```

```{r}
dag.BAH = set.arc(dag.BAH, from = "team", to = "lap_time")
dag.BAH = set.arc(dag.BAH, from = "pitstop_duration", to = "lap_time")
graphviz.plot(dag.BAH)
```
```{r}
ci.test("position", "pitstop_duration", c("lap_time", "previous_position"), test = "mc-jt", data = BAH)
```
```{r}
dag.BAH = set.arc(dag.BAH, from = "pitstop_duration", to = "position")
ci.test("pitstop_duration", "previous_position", test = "mc-jt", data = BAH)
```
```{r}
ci.test("position", "team", c("lap_time", "previous_position"), test = "mi", data = BAH)
```

```{r}
graphviz.plot(dag.BAH)
```


```{r}
BAH.fit = bn.fit(dag.BAH, data = BAH, method = "bayes", iss = 10)
```

```{r}
# We will need the gRain package
library(gRain)
# Use the gRain package to set evidence (condition on an event)
junction = compile(as.grain(BAH.fit))
```

```{r}
jf = setEvidence(junction, names(BAH)[!is.na(BAH[2,])],
                 as.vector(apply(BAH[2,!is.na(BAH[2,])],2,paste)))

graphviz.chart(BAH.fit, grid = TRUE, main = "Original BN")
```

