---
title: "Bayesian Networks in Formula One Race Strategy"
subtitle: "2023 season"
author: "Natalia Kortals"
output: html_document
date: '2024-10-23'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(data.table)
library(stringr)
```

Firstly, we load the data saved into .csv file, acquired using f1dataR package. https://CRAN.R-project.org/package=f1dataR

```{r load all races}
folder_path <- "data/"

file_list <- list.files(path = folder_path, pattern = "2023\\.csv$", full.names = TRUE)
file_list

season_df <- data.frame()
for (file in file_list) {
  temp_df <- read.csv(file)
  # add a variable event that will be useful for filtering                                                                     through races
  temp_df$event <- gsub("2023", "", tools::file_path_sans_ext(basename(file))) 
  season_df <- rbind(season_df, temp_df)
}
```

```{r size of data}
print(dim(season_df))
```

```{r head}
head(season_df)
```
## Summary of the data

Here we can see that we have to discretise and categorise the integer and continuous variables.

```{r summary}
summary(season_df)
```

Next, we plot the variables to understand the dataset: 

```{r laps count plot}
ggplot(season_df, aes(x = event, fill = team)) +
  geom_bar(position = "stack") +
  labs(x = "Event", y = "Count", title = "Sample Size of Laps per Race by Team") +
  scale_fill_brewer(palette = "Paired", name = "Team") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r tyres count plot}
ggplot(season_df, aes(x = event, fill = tyre)) +
  geom_bar(position = "stack") +
  labs(x = "Event", y = "Count", title = "Sample Size of Tyre Compounds per Race") +
  scale_fill_brewer(palette = "Paired", name = "Tyre") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r rainfall count plot}
ggplot(season_df, aes(x = event, fill = rainfall)) +
  geom_bar(position = "stack") +
  labs(x = "Event", y = "Count", title = "Weather Conditions per Race") +
  scale_fill_brewer(palette = "Paired", name = "Rainfall") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r circuit vars heatmap}
# Need a new df with 8 variables, 5 rows, count of unique events with a value

value_counts <- season_df %>%
  pivot_longer(cols = c(traction, asphalt_grip, asphalt_abrasion, track_evol,
                    tyre_stress, braking, lateral, downforce), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n_distinct(event), .groups = "drop") %>%  # Count unique events
  arrange(Variable, Value)

# Print result
print(value_counts)

ggplot(value_counts, aes(x = Variable, y = as.factor(Value), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightyellow", high = "red") +
  labs(title = "Heatmap of Circuit Characteristics Across 22 Races",
       x = "Variable",
       y = "Value",
       fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5))

```


```{r plot event vs laptime}
# if pitstop is na
ggplot(data=subset(season_df, is.na(pitstop_duration)), mapping = aes(x = event, y = lap_time)) +
  geom_boxplot() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(x = "Event", y = "Lap Time", title = "Box Plot of Lap Times at 2023 Races") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```


## Missingness

```{r wrong pitstop data}
# Check how many wrong pitstop_duration values are there and which races they are present in

# Check for time-series-like string entries
time_like <- str_detect(season_df$pitstop_duration, "\\d+:\\d+\\.\\d+")
# Count occurrences
num_time_like <- sum(time_like)
# Check how many time-series-like entries there are per event
miss_pit_per_events <- season_df %>%
  filter(time_like) %>%
  group_by(event) %>%
  summarise(count = n(), .groups = "drop")

print(miss_pit_per_events)

```
It seems that only 7 events are affected by the wrong pitstop data format. Interestingly, for all those laps the lap_time variable is also missing, which seems to be an error or technology failure in collecting the timing data, with the entry being overwritten by a timestamp. 


```{r pitstop count}
# Get total number of pitstops per race to help deciding what to drop
total_pitstops <- season_df %>%
  filter(!is.na(pitstop_duration)) %>%
  group_by(event) %>%
  summarise(count = n(), .groups = "drop")
print(total_pitstops)
```

```{r avg no of pitstops per race}
# get the average of pitstops
mean(total_pitstops$count)
```

In albert_park2023 the incorrect format of pitstop_duration variable accounts for 75% of the pitstop datapoints.
For interlagos2023 and rodriguez2023, the incorrect datapoints are around 25% and 47% of the total pitstops.
In Zaandvort 9 laps are affected, which makes for only around 11% of pitstops at that event.
For races where this occurs only once, such as marina_bay2023, suzuka2023, villeneuve2023 - this mistaken row of data will be dropped.

Setting an allowance of up to 20% pitstop_duration data missing, we disregard data from
albert_park2023, interlagos2023, rodriguez2023 in later analysis. For other events, this rows of data will be dropped as a data cleaning step.


```{r pitstop bins}
# Make a massive loop now

for (round in unique(season_df$event)){
  print(round)
  race_df <- subset(season_df, event ==round)
  # print(head(race_df))

  # # Convert pitstop_duration to numeric
  race_df$pitstop_duration <- as.numeric(race_df$pitstop_duration)
  # Make bins
  race_df$pitstop_duration <- cut(race_df$pitstop_duration, breaks = c(min(race_df$pitstop_duration, na.rm = TRUE), quantile(race_df$pitstop_duration, probs = c(.2, .4, .6, .8),na.rm=TRUE),max(race_df$pitstop_duration, na.rm = TRUE)), ordered_result = TRUE, include.lowest = TRUE, labels = c("fastest_pit", ".2", ".4", ".6", ".8"))
  
  # Add "no_pitstop" to the factor levels for pitstop_duration
  race_df$pitstop_duration <- factor(race_df$pitstop_duration, 
                                  levels = c("no_pitstop", levels(race_df$pitstop_duration)))
  # Replace all NA values with "no_pitstop"
  race_df<- race_df %>% replace_na(list(pitstop_duration = "no_pitstop"))
  
  print(levels(race_df$pitstop_duration))
  
# At the end they have to merge to make a new full season df
  }

```




```{r boxplot pitstop bins vs laptime}
# Boxplot of pitstop bins vs continuous lap_time
ggplot(data=season_df, aes(x=pitstop_duration, y = lap_time)) +
  geom_boxplot()
```




```{r lap time bins}
# Lap time - ordered: the shorter the better
# the fastest lap is included thanks to argument include.lowest = TRUE
season_df$lap_time <- cut(season_df$lap_time, breaks = c(min(season_df$lap_time, na.rm = TRUE), quantile(season_df$lap_time,probs = c(.05, .1, .15, .2, .25, .3,.4,.5,.6, .7,.8, .85, .87, .9, .92, .94, .96, .98, .99),na.rm=TRUE),max(season_df$lap_time, na.rm = TRUE)), include.lowest = TRUE, ordered_result = TRUE, labels = c("fastest", ".05", ".1", ".15", ".2", ".25", ".3",".4",".5",".6", ".7",".8", ".85", ".87", ".9", ".92", ".94", ".96", ".98", ".99"))

ggplot(data=season_df, aes(x=lap_time)) +
  geom_bar()
```


```{r convert to factors and ordinal}

season_df$driver <- factor(season_df$driver)
season_df$team <- factor(season_df$team)

# Circuit characteristics
season_df$traction <- factor(season_df$traction, ordered = TRUE)
season_df$asphalt_grip <- factor(season_df$asphalt_grip, ordered = TRUE)
season_df$asphalt_abrasion <- factor(season_df$asphalt_abrasion, ordered = TRUE)
season_df$track_evol <- factor(season_df$track_evol, ordered = TRUE)
season_df$tyre_stress <- factor(season_df$tyre_stress, ordered = TRUE)
season_df$braking <- factor(season_df$braking, ordered = TRUE)
season_df$lateral <- factor(season_df$lateral, ordered = TRUE)
season_df$downforce <- factor(season_df$downforce, ordered = TRUE)

season_df$tyre <- factor(season_df$tyre) 
# Weather can be ordered too, with Dry being the preferred
season_df$rainfall <- factor(season_df$rainfall, ordered = TRUE)

# Make a variable for position on previous lap
season_df <- season_df %>%
  mutate(previous_position = ifelse(lap_number == 1, start_position, shift(season_df$position, n=1, type="lag")))

# Positions are ordered, quite self-explanatory reason
season_df$previous_position <- factor(season_df$previous_position, ordered = TRUE)
season_df$position <- factor(season_df$position, ordered = TRUE)

# Choosing relevant variables
season_df <- season_df %>%  dplyr::select(c(lap_time, 
                                  tyre, team, 
                                  position, 
                                  previous_position, pitstop_duration
                                  , driver
                                  # , 
                                # traction, asphalt_grip, asphalt_abrasion, 
                                # track_evol, tyre_stress, braking, 
                                # lateral, downforce
                                ))

```



```{r checking NAs}
# check which rows have NAs and why
which(is.na(season_df))
```
```{r deal with NAs}
# # dropping  nas
# season_df <- drop_na(season_df)
```



```{r}
# if starting position is 0, raise a Warning to check if that was a pitlane start
# Group by 'driver' and check for zeros
season_df %>%
  group_by(driver) %>%
  summarize(has_zero = any(previous_position == 0), .groups = "drop") %>%
  filter(has_zero) %>%
  pull(driver) -> drivers_with_zero

# Raise warning if any drivers have zeros
if (length(drivers_with_zero) > 0) {
  warning(paste("The following drivers have zeros in their values:", paste(drivers_with_zero, collapse = ", ")))
}
```



