---
title: "Bayesian Networks in F1: Factors influencing the position in a rac"
subtitle: "2023 season"
author: "Natalia Kortals"
output: html_document
date: '2024-10-23'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(data.table)
```

Firstly, we load the data saved into .csv file, acquired using f1dataR package. https://CRAN.R-project.org/package=f1dataR

```{r load all races}
folder_path <- "data/"

file_list <- list.files(path = folder_path, pattern = "2023\\.csv$", full.names = TRUE)
file_list

combined_df <- data.frame()
for (file in file_list) {
  temp_df <- read.csv(file)
  temp_df$event <- tools::file_path_sans_ext(basename(file)) # add a variable event that will be useful for filtering through races
  combined_df <- rbind(combined_df, temp_df)
}

```
## Missingness

```{r wrong pitstop data}
# Check how many wrong pitstop_duration values are there and which races they are present in

library(stringr)

# Check for time-series-like entries
time_like <- str_detect(combined_df$pitstop_duration, "\\d+:\\d+\\.\\d+")
# Count occurrences
num_time_like <- sum(time_like)

# Check how many time-series-like entries there are per event
unique_events <- combined_df %>%
  filter(time_like) %>%
  group_by(event) %>%
  summarise(count = n(), .groups = "drop")

print(unique_events)

```
It seems that only 7 events are affected by the wrong pitstop data format. Interestingly, for all those laps the lap_time variable is also missing, which seems to be an error in collecting the timing data or those laps being disregarded in some way (unfinished laps). 

In albert_park2023 the incorrect format of pitstop_duration variable accounts for 75% of the pitstop datapoints.
For interlagos2023 and rodriguez2023, the incorrect datapoints are around 25% and 47% of the total pitstops.
In Zaandvort 9 laps are affected, which makes for only around 11% of pitstops at that event.
For races where this occurs only once, such as marina_bay2023, suzuka2023, villeneuve2023 - this mistaken row of data will be dropped.

Setting an allowance of up to 20% pitstop_duration data missing, we disregard data from
albert_park2023, interlagos2023, rodriguez2023 in later analysis. For other events, this rows of data will be dropped as a data cleaning step. 

```{r}
# Get total number of pitstops per race to help deciding what to drop
unique_events2 <- combined_df %>%
  filter(!is.na(pitstop_duration)) %>%
  group_by(event) %>%
  summarise(count = n(), .groups = "drop")
print(unique_events2)
```

```{r}
# get the average of pitstops
mean(unique_events2$count)
```


## Summary of the data

```{r}
summary(combined_df)
```

Here we can see that we have to discretise and categorise the integer and continuous variables.




Next, we plot the variables to understand the dataset: 

```{r plot event vs laptime}
# if pitstop is na
# check for safety car conditions
ggplot(data=subset(combined_df, is.na(pitstop_duration)), mapping = aes(x = event, y = lap_time)) +
  geom_boxplot() + theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r}
# install.packages("GGally")
library(GGally)

ggpairs(combined_df,          # Data frame
        columns = c("traction", "asphalt_grip", "asphalt_abrasion", "track_evol",
                    "tyre_stress", "braking", "lateral", "downforce")) 

```



```{r}
ggplot() + geom_bar(aes(x = combined_df$tyre))
```
```{r laptime race 1}
combined_df %>%
  filter( event == "albert_park2023"  ) %>%
  ggplot( aes(x=lap_time)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)
```
```{r}
combined_df %>%
  filter(lap_number == 1 & (event == "albert_park2023" | event == "hungaroring2023")) %>%
  ggplot() +
    geom_point(aes(x = team, y =lap_time, colour = event))

# ggplot() + geom_point(aes(x = combined_df$driver, y = combined_df$lap_time, colour = combined_df$event))
```


```{r laptime race 2}
combined_df %>%
  filter( event == "miami2023" ) %>%
  ggplot( aes(x=lap_time)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8)
# ggplot() + geom_density(aes(x = combined_df$lap_time))
```


```{r}
max(combined_df$lap_time, na.rm = TRUE)
```

```{r pitstop bins}
# Convert pitstop_duration to numeric
combined_df$pitstop_duration <- as.numeric(combined_df$pitstop_duration)
# Make bins
combined_df$pitstop_duration <- cut(combined_df$pitstop_duration, breaks = c(min(combined_df$pitstop_duration, na.rm = TRUE), quantile(combined_df$pitstop_duration, probs = c(.2, .4, .6, .8),na.rm=TRUE),max(combined_df$pitstop_duration, na.rm = TRUE)), ordered_result = TRUE, include.lowest = TRUE, labels = c("fastest_pit", ".2", ".4", ".6", ".8"))
# Quantiles need more meaningful names

ggplot(data=subset(combined_df, !is.na(pitstop_duration)), aes(x=pitstop_duration)) +
  geom_bar()

# Fill NAs with 0,1 bins
levels(combined_df$pitstop_duration) <- c("fastest_pit", ".2", ".4", ".6", ".8", "no_stop")
combined_df <- combined_df %>% replace_na(list(pitstop_duration = "no_stop")) #change to "NA"

```


```{r}
ggplot(combined_df, aes(x=pitstop_duration)) +
  geom_bar()
```


```{r boxplot pitstop bins vs laptime}
# Boxplot of pitstop bins vs continuous lap_time
ggplot(data=combined_df, aes(x=pitstop_duration, y = lap_time)) +
  geom_boxplot()
```




```{r lap time bins}
# Lap time - ordered: the shorter the better
# the fastest lap is included thanks to argument include.lowest = TRUE
combined_df$lap_time <- cut(combined_df$lap_time, breaks = c(min(combined_df$lap_time, na.rm = TRUE), quantile(combined_df$lap_time,probs = c(.05, .1, .15, .2, .25, .3,.4,.5,.6, .7,.8, .85, .87, .9, .92, .94, .96, .98, .99),na.rm=TRUE),max(combined_df$lap_time, na.rm = TRUE)), include.lowest = TRUE, ordered_result = TRUE, labels = c("fastest", ".05", ".1", ".15", ".2", ".25", ".3",".4",".5",".6", ".7",".8", ".85", ".87", ".9", ".92", ".94", ".96", ".98", ".99"))

ggplot(data=combined_df, aes(x=lap_time)) +
  geom_bar()
```


```{r convert to factors and ordinal}

combined_df$driver <- factor(combined_df$driver)
combined_df$team <- factor(combined_df$team)

# Circuit characteristics
combined_df$traction <- factor(combined_df$traction, ordered = TRUE)
combined_df$asphalt_grip <- factor(combined_df$asphalt_grip, ordered = TRUE)
combined_df$asphalt_abrasion <- factor(combined_df$asphalt_abrasion, ordered = TRUE)
combined_df$track_evol <- factor(combined_df$track_evol, ordered = TRUE)
combined_df$tyre_stress <- factor(combined_df$tyre_stress, ordered = TRUE)
combined_df$braking <- factor(combined_df$braking, ordered = TRUE)
combined_df$lateral <- factor(combined_df$lateral, ordered = TRUE)
combined_df$downforce <- factor(combined_df$downforce, ordered = TRUE)

combined_df$tyre <- factor(combined_df$tyre, ordered = TRUE, levels = c("SOFT", "MEDIUM", "HARD")) 
# Weather can be ordered too, with Dry being the preferred
combined_df$rainfall <- factor(combined_df$rainfall, ordered = TRUE)

# Make a variable for position on previous lap
combined_df <- combined_df %>%
  mutate(previous_position = ifelse(lap_number == 1, start_position, shift(combined_df$position, n=1, type="lag")))

# Positions are ordered, quite self-explanatory reason
combined_df$previous_position <- factor(combined_df$previous_position, ordered = TRUE)
combined_df$position <- factor(combined_df$position, ordered = TRUE)

# Choosing relevant variables
combined_df <- combined_df %>%  dplyr::select(c(lap_time, 
                                  tyre, team, 
                                  position, 
                                  previous_position, pitstop_duration
                                  , driver
                                  # , 
                                # traction, asphalt_grip, asphalt_abrasion, 
                                # track_evol, tyre_stress, braking, 
                                # lateral, downforce
                                ))

```



```{r checking NAs}
# check which rows have NAs and why
which(is.na(combined_df))
```
```{r deal with NAs}
# # dropping  nas
# combined_df <- drop_na(combined_df)
```


Plots 

```{r}
ggplot() + geom_point(aes(x = combined_df$pitstop_duration, y = combined_df$lap_time))
```
```{r}
ggplot() + geom_point(aes(x = combined_df$team, y = combined_df$pitstop_duration))
```

```{r}
# if starting position is 0, raise a Warning to check if that was a pitlane start
# Group by 'driver' and check for zeros
combined_df %>%
  group_by(driver) %>%
  summarize(has_zero = any(previous_position == 0), .groups = "drop") %>%
  filter(has_zero) %>%
  pull(driver) -> drivers_with_zero

# Raise warning if any drivers have zeros
if (length(drivers_with_zero) > 0) {
  warning(paste("The following drivers have zeros in their values:", paste(drivers_with_zero, collapse = ", ")))
}
```

## Starting with a know DAG.

```{r}
library(bnlearn)
library(Rgraphviz)
dag = empty.graph(nodes = c("lap_time", "tyre", "team", "position", 
                      "previous_position", "pitstop_duration", "driver")) # "rainfall",
arc.set = matrix(c("team", "position",
                   "driver", "position",
                   "lap_time", "position",
                   "tyre", "lap_time",
                  "pitstop_duration", "lap_time",
                   "pitstop_duration", "position",
                  # "rainfall", "lap_time",
                  # "rainfall", "tyre",
                  "team", "pitstop_duration",
                  "previous_position", "position"
                  
                  ), 
                 byrow = TRUE, ncol=2, dimnames =list(NULL,c("from","to")))
arcs(dag) = arc.set
graphviz.plot(dag)
```
```{r}
arc.strength(dag, data = combined_df, criterion = "x2")
```


```{r}
bn.bayes = bn.fit(dag,data=combined_df, method = "bayes", iss = 10)

bn.bayes$pitstop_duration
```
```{r}
bn.bayes$lap_time
```

```{r}
bn.bayes$tyre
```
## DAG uknown case

Learning DAG using hill climbing and  blacklists

```{r}
# Blacklist variables so that chronological order is kept
bl = matrix(c("position", "previous_position",
              "pitstop_duration", "tyre",
              "tyre", "pitstop_duration",
              "lap_time", "tyre"
              
              ), 
            ncol = 2, byrow = TRUE)

# Nothing from position
for (name in names(combined_df)){
  bl2 = matrix(c("position", name), ncol=2,
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl2)
}


# Nothing to previous position apart from team
for (name in names(combined_df)){
  if (name != "team"){
  bl3 = matrix(c(name, "previous_position"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl3)}
}

# Nothing to team
for (name in names(combined_df)){
  bl4 = matrix(c(name, "team"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl4)
}

bl
```


```{r}
unknown_dag = empty.graph(nodes = names(combined_df))
wl = matrix(c("previous_position", "position",
              "lap_time", "position"
              ), 
            ncol = 2, byrow = TRUE)
arcs(unknown_dag) = wl
graphviz.plot(unknown_dag)

```


```{r}
# Use hill-climbing to determine the DAG
combined_df.hc =hc(combined_df, blacklist = bl, whitelist = wl)
graphviz.plot(combined_df.hc)
```

```{r}
arc.strength(combined_df.hc, data = combined_df, criterion = "x2")
```

```{r}
score(dag, data = combined_df, type = "bic")
score(combined_df.hc, data = combined_df, type = "bic")
```
```{r}
score(dag, data = combined_df, type = "bde")
score(combined_df.hc, data = combined_df, type = "bde")
```

```{r}
# Alt hill cclimbing with no rules
altcombined_df.hc =hc(combined_df, blacklist = bl)
score(altcombined_df.hc, data = combined_df, type = "bde")
graphviz.plot(altcombined_df.hc)
```

## Use incremental association Markov blanket to learn the DAG

```{r}
combined_df.iamb =iamb(combined_df, blacklist = bl, whitelist = wl, test ='mi')
graphviz.plot(combined_df.iamb)
```


## Perform independence tests
If the p-value is very small, there is no evidence for independence so we add the arc to the dag.
```{r}
# Testing for conditional independence: position
ci.test("team", "previous_position", test = "mi", data = combined_df)
```
```{r}
dag.combined_df = set.arc(combined_df.iamb, from = "team", to = "previous_position")

ci.test("team", "pitstop_duration", test = "mi", data = combined_df)
```
```{r}
ci.test("lap_time","pitstop_duration", test = "jt", data = combined_df)
```

```{r}
dag.combined_df = set.arc(dag.combined_df, from = "pitstop_duration", to = "lap_time")
graphviz.plot(dag.combined_df)
```

```{r}
ci.test("lap_time", "tyre", "pitstop_duration", test = "jt", data = combined_df)
```

```{r}
ci.test("lap_time", "team", "pitstop_duration", test = "mi", data = combined_df)
```

```{r}

```
```{r}

```

