---
title: "Bayesian Networks in Formula One Race Strategy"
subtitle: "2023 season"
author: "Natalia Kortals"
output: html_document
date: '2024-10-23'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(data.table)
library(stringr)
library(bnlearn)
library(Rgraphviz)
library(GGally)
```

Firstly, we load the data saved into .csv file, acquired using f1dataR package. https://CRAN.R-project.org/package=f1dataR

```{r load all races}
folder_path <- "data/"

file_list <- list.files(path = folder_path, pattern = "2023\\.csv$", full.names = TRUE)
file_list

season_df <- data.frame()
for (file in file_list) {
  temp_df <- read.csv(file)
  # add a variable event that will be useful for filtering                                                                     through races
  temp_df$event <- gsub("2023", "", tools::file_path_sans_ext(basename(file))) 
  season_df <- rbind(season_df, temp_df)
}
```

```{r size of data}
print(dim(season_df))
```

```{r head}
head(season_df)
```
## Summary of the data

Here we can see that we have to discretise and categorise the integer and continuous variables.

```{r summary}
summary(season_df)
```

Next, we plot the variables to understand the dataset: 

```{r laps count plot}
ggplot(season_df, aes(x = event, fill = team)) +
  geom_bar(position = "stack") +
  labs(x = "Event", y = "Count", title = "Sample Size of Laps per Race by Team") +
  scale_fill_brewer(palette = "Paired", name = "Team") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r tyres count plot}
ggplot(season_df, aes(x = event, fill = tyre)) +
  geom_bar(position = "stack") +
  labs(x = "Event", y = "Count", title = "Sample Size of Tyre Compounds per Race") +
  scale_fill_brewer(palette = "Paired", name = "Tyre") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r rainfall count plot}
ggplot(season_df, aes(x = event, fill = rainfall)) +
  geom_bar(position = "stack") +
  labs(x = "Event", y = "Count", title = "Weather Conditions per Race") +
  scale_fill_brewer(palette = "Paired", name = "Rainfall") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r circuit vars heatmap}
# Need a new df with 8 variables, 5 rows, count of unique events with a value

value_counts <- season_df %>%
  pivot_longer(cols = c(traction, asphalt_grip, asphalt_abrasion, track_evol,
                    tyre_stress, braking, lateral, downforce), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n_distinct(event), .groups = "drop") %>%  # Count unique events
  arrange(Variable, Value)

# Print result
print(value_counts)

ggplot(value_counts, aes(x = Variable, y = as.factor(Value), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightyellow", high = "red") +
  labs(title = "Heatmap of Circuit Characteristics Across 22 Races",
       x = "Variable",
       y = "Value",
       fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5))

```
```{r plot team vs position}
# if pitstop is na
ggplot(data=season_df, mapping = aes(x = team, y = start_position)) +
  geom_boxplot() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(x = "Team", y = "Starting Position", title = "Box Plot of Starting Positions by Team in 2023") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r plot event vs laptime}
# if pitstop is na
ggplot(data=subset(season_df, is.na(pitstop_duration)), mapping = aes(x = event, y = lap_time)) +
  geom_boxplot() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(x = "Event", y = "Lap Time", title = "Box Plot of Lap Times at 2023 Races") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r plot position vs laptime}
# if pitstop is na
ggplot(data=subset(season_df, is.na(pitstop_duration)), mapping = aes(x = lap_time, y = position)) +
  geom_jitter() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(x = "Lap Time", y = "Position", title = "Lap Time vs Position at 2023 Races") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r pairplot}
ggpairs(season_df, columns = c("lap_time", "position"), columnLabels = c("Lap Time", "Position")) 
```


## Missingness

```{r wrong pitstop data}
# Check how many wrong pitstop_duration values are there and which races they are present in

# Check for time-series-like string entries
time_like <- str_detect(season_df$pitstop_duration, "\\d+:\\d+\\.\\d+")
# Count occurrences
num_time_like <- sum(time_like)
# Check how many time-series-like entries there are per event
miss_pit_per_events <- season_df %>%
  filter(time_like) %>%
  group_by(event) %>%
  summarise(count = n(), .groups = "drop")

print(miss_pit_per_events)

```
It seems that only 7 events are affected by the wrong pitstop data format. Interestingly, for all those laps the lap_time variable is also missing, which seems to be an error or technology failure in collecting the timing data, with the entry being overwritten by a timestamp. 


```{r pitstop count}
# Get total number of pitstops per race to help deciding what to drop
total_pitstops <- season_df %>%
  filter(!is.na(pitstop_duration)) %>%
  group_by(event) %>%
  summarise(count = n(), .groups = "drop")
print(total_pitstops)
```

```{r avg no of pitstops per race}
# get the average of pitstops
mean(total_pitstops$count)
```

In albert_park the incorrect format of pitstop_duration variable accounts for 75% of the pitstop datapoints.
For interlagos and rodriguez, the incorrect datapoints are around 25% and 47% of the total pitstops.
In Zaandvort 9 laps are affected, which makes for only around 11% of pitstops at that event.
For races where this occurs only once, such as marina_bay, suzuka, villeneuve - this mistaken row of data will be dropped.

Setting an allowance of up to 20% pitstop_duration data missing, we disregard data from
albert_park, interlagos, rodriguez in later analysis. For other events, this rows of data will be dropped as a data cleaning step.

```{r}
# # remove albert_park, interlagos, rodriguez
# season_df_filtered <- season_df %>% filter(!event %in% c("albert_park", "interlagos", "rodriguez"))

#find out the percentage of missing lap time data
sum(is.na(season_df$lap_time))/dim(season_df)[1]

#drop the missing lap_times as they account for less than 2% of the data
season_df_filtered <- season_df %>% filter(!is.na(lap_time))

# filter out Alpine at Hungaroring (only 1 or 2 laps per driver due to crash)
season_df_filtered <- season_df_filtered %>% filter(event != "hungaroring" | team != "Alpine")

```


```{r}
# list of levels of previous position
prev_pos_levels <- c("1",  "2" , "3", "4", "5" , "6",  "7",  "8",  "9",  "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "0")
pos_levels <-  c("1",  "2" , "3", "4", "5" , "6",  "7",  "8",  "9",  "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20")
```


```{r}
# Create an empty list to store each processed round
race_list <- list()

for (round in unique(season_df_filtered$event)) {
  print(round)
  race_df <- subset(season_df_filtered, event == round)

  # Convert pitstop_duration to numeric
  race_df$pitstop_duration <- as.numeric(race_df$pitstop_duration)
  
  # Make bins
  race_df$pitstop_duration <- cut(
    race_df$pitstop_duration, 
    breaks = c(
      min(race_df$pitstop_duration, na.rm = TRUE), 
      quantile(race_df$pitstop_duration, probs = c(.2, .4, .6, .8), na.rm=TRUE), 
      max(race_df$pitstop_duration, na.rm = TRUE)
    ), 
    ordered_result = TRUE, 
    include.lowest = TRUE, 
    labels = c("fastest_pit", "20", "40", "60", "80")
  )
  
  # Add "no_pitstop" to the factor levels for pitstop_duration
  race_df$pitstop_duration <- factor(
    race_df$pitstop_duration, 
    levels = c("no_pitstop", levels(race_df$pitstop_duration))
  )
  
  # Replace all NA values with "no_pitstop"
  race_df <- race_df %>% replace_na(list(pitstop_duration = "no_pitstop"))
  
  # Lap time bins
  race_df$lap_time <- cut(race_df$lap_time, 
                            breaks = c(min(race_df$lap_time, na.rm = TRUE), 
                            quantile(race_df$lap_time,probs = c(.02, .05, .1, .15, .2, .3,.4,.5,.6, .7,.8, .85, .9, .95, .98),na.rm=TRUE), 
                            max(race_df$lap_time, na.rm = TRUE)), include.lowest = TRUE, ordered_result = TRUE, labels = c("fastest", "02", "05", "10", "15", "20", "30","40","50","60", "70","80", "85", "90", "95", "98"))
  
  # Convert to factors or ordered factors
  race_df$driver <- factor(race_df$driver)
  race_df$team <- factor(race_df$team)
  race_df$tyre <- factor(race_df$tyre) 
  race_df$rainfall <- factor(race_df$rainfall)
  
  
  # Make a variable for position on previous lap
  race_df <- race_df %>%
    mutate(previous_position = ifelse(lap_number == 1, start_position, shift(race_df$position, n=1, type="lag")))
  
  # Positions are ordered, quite self-explanatory reason
  race_df$previous_position <- factor(race_df$previous_position, ordered = TRUE, levels = prev_pos_levels)
  race_df$position <- factor(race_df$position, ordered = TRUE, levels = pos_levels)

  
  # Store the modified race_df in the list
  race_list[[round]] <- race_df
}

# Merge all processed race_df dataframes into a new full-season dataframe
full_season_df <- bind_rows(race_list)

```



```{r convert circuit characteristics to factors and ordinal}

# Circuit characteristics
full_season_df$traction <- factor(full_season_df$traction, ordered = TRUE)
full_season_df$asphalt_grip <- factor(full_season_df$asphalt_grip, ordered = TRUE)
full_season_df$asphalt_abrasion <- factor(full_season_df$asphalt_abrasion, ordered = TRUE)
full_season_df$track_evol <- factor(full_season_df$track_evol, ordered = TRUE)
full_season_df$tyre_stress <- factor(full_season_df$tyre_stress, ordered = TRUE)
full_season_df$braking <- factor(full_season_df$braking, ordered = TRUE)
full_season_df$lateral <- factor(full_season_df$lateral, ordered = TRUE)
full_season_df$downforce <- factor(full_season_df$downforce, ordered = TRUE)

#Save each of the races to be removed as seperate dfs

interlagos_df <- full_season_df %>% filter(event =="interlagos")

albert_park_df <- full_season_df %>% filter(event =="albert_park")

rodriguez_df <- full_season_df %>% filter(event =="rodriguez")


# remove albert_park, interlagos, rodriguez
full_season_df <- full_season_df %>% filter(!event %in% c("albert_park", "interlagos", "rodriguez"))


# Choosing relevant variables
season23_full <- full_season_df %>%  dplyr::select(c( driver, lap_time, 
                                tyre, team, 
                                position, rainfall,
                                previous_position, pitstop_duration,
                                traction, asphalt_grip, asphalt_abrasion,
                                track_evol, tyre_stress, braking,
                                lateral, downforce
                                ))

```

```{r corr of discretised LT vs position}
library(polycor)
polychor(season23_full$lap_time, season23_full$position)
```

```{r}
# Show rows where starting position is 0, to check if that was a pitlane start
full_season_df %>%
  group_by(event, driver) %>%
  filter(previous_position == 0) %>%
  print(full_season_df)
```
# Structure learning


## Blacklists and whitelists
Make a blacklist representing the natural temporal ordering of variables

```{r}
# Blacklist variables so that chronological order is kept
big_bl = matrix(c(
              "lap_time", "pitstop_duration",
              "lap_time", "tyre"
             ), 
            ncol = 2, byrow = TRUE)

# Nothing from position
for (name in names(season23_full)){
  bl2 = matrix(c("position", name), ncol=2,
            dimnames =list(NULL,c("from","to")))
  big_bl = rbind(big_bl, bl2)
}


# Nothing to team
for (name in names(season23_full)){
  bl3 = matrix(c(name, "team"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  big_bl = rbind(big_bl, bl3)
}

# Nothing to rainfall
for (name in names(season23_full)){
  bl3 = matrix(c(name, "rainfall"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  big_bl = rbind(big_bl, bl3)
}

# Nothing to circuit characteristics from other variables
circuit_chars <- c("traction", "asphalt_grip", "asphalt_abrasion",
                                "track_evol", "tyre_stress", "braking",
                                "lateral", "downforce")
other_vars <- c("driver", "lap_time", 
                                "tyre", "team", "rainfall",
                                "previous_position", "pitstop_duration")
for (name in names(season23_full)){
  for (circ in circuit_chars){
    bl4 = matrix(c(name, circ), ncol=2, 
              dimnames =list(NULL,c("from","to")))
    big_bl = rbind(big_bl, bl4)
}}

big_bl

# Whitelist
wl = matrix(c(
              "previous_position", "position"
             ), 
            ncol = 2, byrow = TRUE)

```

## IAMB


```{r}
season23_full.iamb =iamb(season23_full, blacklist = big_bl, whitelist = wl, test ='mi')
highlight = list(nodes = circuit_chars, col = "blue", fill = "lightblue")
graphviz.plot(season23_full.iamb, render = TRUE, highlight = highlight)
```
```{r}
# Choosing relevant variables
season23 <- full_season_df %>%  dplyr::select(c( driver, lap_time, 
                                tyre, team, 
                                position, rainfall,
                                previous_position, pitstop_duration,
                                traction, tyre_stress 
                                ))

# New blacklist variables so that chronological order is kept
bl = matrix(c(
              "lap_time", "pitstop_duration",
              "lap_time", "tyre"
             ), 
            ncol = 2, byrow = TRUE)

# Nothing from position
for (name in names(season23)){
  bl2 = matrix(c("position", name), ncol=2,
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl2)
}


# Nothing to team
for (name in names(season23)){
  bl3 = matrix(c(name, "team"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl3)
}

# Nothing to rainfall
for (name in names(season23)){
  bl3 = matrix(c(name, "rainfall"), ncol=2, 
            dimnames =list(NULL,c("from","to")))
  bl = rbind(bl, bl3)
}

# Nothing to circuit characteristics from other variables
circuit_chars_select <- c("traction", "tyre_stress")

for (name in names(season23)){
  for (circ in circuit_chars_select){
    bl4 = matrix(c(name, circ), ncol=2, 
              dimnames =list(NULL,c("from","to")))
    bl = rbind(bl, bl4)
}}

bl

# Whitelist
wl = matrix(c(
              "previous_position", "position"
             ), 
            ncol = 2, byrow = TRUE)

```

```{r}
season23.iamb =iamb(season23, blacklist = bl, whitelist = wl, test ='mi')
graphviz.plot(season23.iamb)
score(season23.iamb, data = season23, type = "bic")
```

## Hill-climbing

Learning DAG using hill climbing and whitelists, blacklists
```{r}
# Use hill-climbing to determine the DAG with blacklisting
season23_full.hc =hc(season23_full, blacklist = big_bl, whitelist = wl)

highlight = list(nodes = circuit_chars, col = "blue", fill = "lightblue")
graphviz.plot(season23_full.hc, render = TRUE, highlight = highlight)
score(season23_full.hc, data = season23_full, type = "bic")
```



```{r}
# Use hill-climbing to determine the DAG with blacklisting
season23.hc =hc(season23, blacklist = bl, whitelist = wl)
graphviz.plot(season23.hc)
score(season23.hc, data = season23, type = "bic")
```



Perform independence tests:

```{r}
# Does driver influence previous position?
ci.test("previous_position","driver", test = "mc-mi", data = season23)
```
```{r}
# Does pitstop_duration influence lap_time?
ci.test("lap_time", "pitstop_duration", test = "mc-jt", data = season23)
```
```{r}
# Does pitstop_duration influence lap_time given tyre?
ci.test("lap_time", "pitstop_duration", "tyre", test = "mc-mi", data = season23)
```

```{r}
# Does tyre influence lap_time?
ci.test("lap_time","tyre", "pitstop_duration", test = "mc-mi", data = season23)
```

```{r}
# Does rainfall influence tyre?
ci.test("tyre", "rainfall", c("traction", "tyre_stress"), test = "mc-mi", data = season23)
```
```{r}
# Does lap_time influence position?
ci.test("position", "lap_time", "previous_position", test = "mc-jt", data = season23)
```


```{r}
tested_season23.hc = set.arc(season23.hc, from = "tyre", to = "lap_time")
tested_season23.hc = set.arc(tested_season23.hc, from = "lap_time", to = "position")
graphviz.plot(tested_season23.hc)
```
```{r}
score(tested_season23.hc, data = season23, type = "bic")
```


```{r}
arc.strength(tested_season23.hc, data = season23, criterion = "bic")
```
```{r}
arc.strength(tested_season23.hc, data = season23, criterion = "mi")
```

## Prediction
Use the deleted races for prediction (interlagos example)


```{r}
# Choosing relevant variables
int_pred <- interlagos_df %>%  dplyr::select(c( driver, lap_time, 
                                tyre, team, 
                                # position,
                                rainfall,
                                previous_position, pitstop_duration,
                                traction, tyre_stress
                                ))

# Ensure the levels match (some drivers were changed during the season)
levels(int_pred$driver) <- levels(season23$driver)

# Predict
season23.fit = bn.fit(tested_season23.hc, data = season23)
int.prediction = predict(season23.fit, data= int_pred, node = "position", method = "bayes-lw")
int.prediction
```
```{r}
# Assess the predicitons
int.predicted.pos = predict(season23.fit, data= int_pred, node = "position", method = "bayes-lw")

predicted_values_numeric <- as.numeric(as.character(int.predicted.pos))
true_values_numeric <- as.numeric(as.character(interlagos_df$position))

mae <- mean(abs(predicted_values_numeric - true_values_numeric), na.rm = TRUE)
print(paste("MAE:", round(mae, 2)))

rmse <- sqrt(mean((predicted_values_numeric - true_values_numeric)^2, na.rm = TRUE))
print(paste("RMSE:", round(rmse, 2)))


```
Barcelona what-if prediction
```{r}
# Choosing relevant variables
bar_pred <- full_season_df %>% filter(event == "catalunya") %>%
  dplyr::select(c( driver, lap_time, 
                                tyre, team, 
                                # position,
                                rainfall,
                                previous_position, pitstop_duration,
                                traction, tyre_stress
                                ))

# Ensure the levels match (some drivers were changed during the season)
levels(bar_pred$driver) <- levels(season23$driver)

# Predict
season23.fit = bn.fit(tested_season23.hc, data = season23)
bar.prediction = predict(season23.fit, data= bar_pred, node = "position", method = "bayes-lw")
bar.prediction
```


